---
title: "PS5 Template"
author: "Phu Quan"
date: "2/13/2023"
output: html_document

git-hub-link : https://github.com/Quann02/PS5-Assisgnment.git
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Disclaimer: this serves as a template for you to compose your answer for PS4. It is 100% hardcoded, so error may exists. Please refer to the official spec for a full list of questions and requirements.

# Instructions

This is a problem set about rmarkdown and plotting (using ggplot). Unlike the previous problem sets, this one does not give you a ready-made GH repo with a code file--it is now your task to create a repo and include your rmarkdown file in there.\

You should answer the questions below in that file, knit it, and submit both the compiled html and link to your repo on canvas.\

-   This problem sets asks you to write extensively when commenting your results. Please write clearly! Answer questions in a way that if the code chunks are hidden then the result is still readable!\
    -   All substantial questions need explanations. You do not have to explain the simple things like "how many rows are there in data", but if you make a plot of life expectancy, then you should explain what does the plot tell you.\
    -   Write explanations as markdown and use the styles like bold and italic as appropriate.\
-   Do not print too much results. It is all well to print a few lines of data for evaluation/demonstration purposes. But do not print dozens (or thousands!) of lines--no one bothers to look at that many numbers. You will lose points for annoying others (here your graders, but later potentially your boss).\
-   Do not make code lines too long. 80-100 characters is a good choice. Your grader may not be able to follow all the code if the line is too long--most of us are using small laptop screens! (And again--you want to keep your graders happy!)\

# Gapminder data

We use gapminder dataset, downloaded from [**https://www.gapminder.org/data/**](https://www.gapminder.org/data/), however, the data structure there is quire complex, please use the dataset provided on canvas (in files/data).\
The variables are:\
**name** country name\
**iso3** 3-letter country code\
**iso2** 2-letter country code\
**region** broad geographic region\
**sub-region** more precise region\
**intermediate-region**\
**time** year\
**totalPopulation** total population\
**GDP_PC** GDP per capita (constant 2010 US\$)\
**accessElectricity** Access to electricity (% of population)\
**agriculturalLand** Agricultural land (sq. km)\
**agricultureTractors** Agricultural machinery, tractors (count)\
**cerealProduction** Cereal production (metric tons)\
**feritilizerHa** Fertilizer consumption (kilograms per hectare of arable land)\
**fertilityRate** total fertility rate (births per woman)\
**lifeExpectancy** Life expectancy at birth, total (years)\
**childMortality** Mortality rate, under-5 (per 1,000 live births)\
**youthFemaleLiteracy** Literacy rate, youth female (% of females ages 15-24)\
**youthMaleLiteracy** Literacy rate, youth male (% of males ages 15-24)\
**adultLiteracy** Literacy rate, adult total (% of people ages 15 and above)\
**co2** CO2 emissions (kt)\
**greenhouseGases** Total greenhouse gas emissions (kt of CO2 equivalent)\
**co2_PC** CO2 emissions (metric tons per capita)\
**pm2.5_35** PM2.5 pollution, population exposed to levels exceeding WHO Interim Target-1 value 36ug/m3\
**battleDeaths** Battle-related deaths (number of people)\

# 1 Load and check data (5pt)

You first task is to do a very simple data check:

1.  (1pt) For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title. See [**https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating**](https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating).\

2.  (2pt) Load data. How many rows/columns do we have?\

```{r}
library(tidyverse)

getwd()
setwd("/Users/phu/Documents/info201_assignment/PS5-Assisgnment")
df <- read_delim("gapminder.csv.bz2")

dim(df)
```

3.  (2pt) Print a small sample of data. Does it look OK?

```{r}

head(df)

```

------------------------------------------------------------------------

# 2 Descriptive statistics (15pt)

1.  (3pt) How many countries are there in the dataset? Analyze all three: *iso3*, *iso2* and *name*.

```{r}
## View(df)
unique_iso2 <- distinct(df, iso2)
count(unique_iso2)

unique_iso3 <- distinct(df, iso3)
count(unique_iso3)

unique_countries <- distinct(df, name)
count(unique_countries)
```

2.  If you did this correctly, you saw that there are more iso-2 codes than names, and there are even more *iso3*-codes. What is going on? Can you find it out?

#### The reason for this is because iso3 has 4 NA.

```{r}
df %>% 
  group_by(name) %>% 
  summarise(n_iso3 = n_distinct(iso3))
filter(count_iso3, n_iso3 > 1)

```

    (a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?

#### *There are no countries that has more than one name, but there are duplicates of NA.*

```{r}
df %>%
  group_by(iso2) %>%
  summarize(n = n_distinct(name)) %>%
  arrange(desc(n)) %>%
  select(iso2, n) %>%
  print()
```

    (b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have more than one iso3-code? What are these countries? Hint: two of these entitites are *CHANISL* and *NLD CURACAO*.

#### *Yes there were 4 countries names with more than on iso3-code. It was Chanisl, GBM, KOS and NLD_CURACAO.*

```{r}
df %>% 
   group_by(name) %>% 
   summarise(ncount = n_distinct(iso3)) %>%
   select(name,ncount) %>%
   arrange(desc(ncount)) %>%
   print()

df %>% 
   filter(is.na(name)) %>% 
   distinct(iso3)

```

3.  (2pt) What is the minimum and maximum year in these data?

```{r}
df %>% 
  filter(!is.na(time)) %>% 
  summarise(maximum_year = max(time), minimum_year = min(time))
```

------------------------------------------------------------------------

# 3 CO2 emissions (30pt)

Next, let's analyze CO2 emissions.

1.  (2pt) How many missing co2 emissions are there for each year? Analyze both missing *CO2* and *co2_PC*. Which years have most missing data?\

#### *In the year 2017, 2018 and 2018. It has the most missing data.*

```{r}

df %>%
    filter(is.na(co2_PC),is.na(co2)) %>% 
    group_by(time) %>% 
    summarise(years= n()) %>%
    arrange(years) %>% 
    tail(3)
```

2.  (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see.\

#### In the plot below it shows that China has emitted the most CO2 compare to any other countries, while Afghanistan has emitted the least CO2 compare to any other countries. For United State we can see that it is decreasing it's emission. Lastly the rest are just flat in CO2 emission.

```{r}

countries <- c("United States of America", "China", "India", "Afghanistan", "United Arab Emirates", "Australia")

co2_by_country <- df %>%
  filter(name %in% countries) %>%
  group_by(name, time) %>%
  summarize(sum_co2 = sum(co2))

ggplot(data = co2_by_country, aes(x = time, y = sum_co2, color = name)) +
  geom_line() +
  labs(title = "Total CO2 Emissions over Time",
       x = "Year",
       y = "Total CO2 Emissions (kilotons)",
       color = "Country")

```

3.  (5pt) Now let's analyze the CO2 emissions per capita (*co2_PC*). Make a similar plot of thesame countries. What does this figure suggest?\

#### In this plot every countries but United Arab has a flat Co2 emission. The reason why United Arab might have large increase of emission and then decrease is because it is probably becoming a more develop country or it enter the industrial revolution.

```{r}
countries <- c("United States of America", "China", "India", "Afghanistan", "United Arab Emirates", "Australia")

co2_by_country <- df %>%
  filter(name %in% countries) %>%
  group_by(name, time) %>%
  summarize(co2PC_sum = sum(co2_PC))

ggplot(data = co2_by_country, aes(x = time, y = co2PC_sum, color = name)) +
  geom_line() +
  labs(title = "Total CO2 Emissions",
       x = "Years",
       y = "Total CO2 Emissions",
       color = "Country")
```

4.  (6pt) Compute average CO2 emissions per capita across the continents (assume *region* is the same as continent). Comment what do you see.\
    Note: just compute averages over countries and ignore the fact that countries are of different size.\
    Hint: Americas 2016 should be 4.80.

#### *Europe has the highest average co2 emissions per capital, while Africa has the lowest average co2 emission per capital.*

```{r}
df %>% 
   filter(time == "2016",!is.na(co2_PC)) %>% 
   group_by(region) %>% 
   summarise(avgC02 = mean(co2_PC))
```

5.  (7pt) Make a barplot where you show the previous results--average CO2 emissions per capita across continents in 1960 and 2016.\
    Hint: it should look something along these lines(**Please refer to the spec for the graph**)\

#### Looking at the bar graph, America has decrease there average co2 emission per capita from the year 1960 to 2016. However Asia, Europe, Africa and Oceania has increase there average co2 emission through 1960 and 2016.

```{r}
years <- c(1960, 2016)

co2_result <- df %>%
  filter(time %in% years & !is.na(co2_PC), !is.na(region)) %>%
  group_by(region, time) %>%
  summarize(avg_co2_PC = mean(co2_PC)) 

ggplot(co2_result, aes(x = region, y = avg_co2_PC, fill = as.factor(time))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Region", y = "Average CO2 emissions per capita", fill = "Years") +
  ggtitle("1960 and 2016 average co2 per cap")
```

6.  Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume *region* is continent).\

#### The largest CO2 emitter for Africa is Angola, Burundi and Benin. While the smallest is South Africa, Zambia and Zimbaebwe.

```{r}
region_name <- df %>%
  filter(!is.na(region)) %>%
  group_by(region) %>%
  distinct(region)
print(region_name)

df %>% 
    filter(time == "2019") %>% 
    group_by(region) %>% 
    arrange(co2_PC) %>% 
    slice_head(n = 3) %>% 
    select(region, name)

df %>% 
    filter(time == "2019") %>% 
    group_by(region) %>% 
    arrange(co2_PC) %>% 
    slice_tail(n = 3) %>% 
    select(region, name)
```

------------------------------------------------------------------------

# 4 GDP per capita (50pt)

Let's look at GDP per capita (*GDP_PC*)

1.  (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960. Make the point size dependent on the country size, and color those according to the continent. Feel free to adjust the plot in other ways to make it better.\
    Comment what do you see there.

#### Europe has the highest LE per capita compare to any other countries, while Africa has the lowest LE per capita compare to any other countries.

```{r}
df_1960 <- df %>% 
  filter(time == "1960", !is.na(GDP_PC)) %>%
  group_by(region)

ggplot(df_1960, aes(x = GDP_PC, y = lifeExpectancy, color = region, size = totalPopulation)) +
  geom_point(alpha = 0.7) +
  scale_size(range = c(1, 20)) +
  labs(title = "GDP per capita vs Life Expectancy by Country in 1960",
       x = "GDP per capita",
       y = "Life Expectancy",
       color = "Region",
       size = "population") +
  theme_bw()


```

2.  (4pt) Make a similar plot, but this time use 2019 data only.\

#### *Compare to the 2016 plot, the 2019 plots shows that more countries has increase there life expectancy. In addition to that, countries has increase there population LE compare to 2016.*

```{r}

df_1960 <- df %>% 
  filter(time == "2019", !is.na(GDP_PC), !is.na(region)) %>%
  group_by(region)

ggplot(df_1960, aes(x = GDP_PC, y = lifeExpectancy, color = region, size = totalPopulation)) +
  geom_point(alpha = 0.7) +
  scale_size(range = c(1, 20)) +
  labs(title = "GDP per capita vs Life Expectancy by Country in 2016",
       x = "GDP per capita",
       y = "Life Expectancy",
       color = "Region",
       size = "population") +
  theme_bw()
```

3.  (6pt) Compare these two plots and comment what do you see. How has world developed through the last 60 years?

#### When comparing this two plots, I see that every continent but Africa has a dramatic increase of life expectancy per capital. This was proably due to that fact that the world has develope better techology and medicine for people to live longer over the last 60 years.

4.  (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures?\
    Note: here as *average*I mean just average over countries, ignore the fact that countries are of different size.

#### For all the regions you can see that the LE of each continent improve from the 1960 to 2019.

```{r}
LE_Mean_1960 <- df %>%
  group_by(region) %>%
  filter(time == "1960", !is.na(lifeExpectancy)) %>%
  summarize(avg_LE = mean(lifeExpectancy))
print(LE_Mean_1960)

LE_Mean_2019 <- df %>%
  group_by(region) %>%
  filter(time == "2019", !is.na(lifeExpectancy)) %>%
  summarize(avg_le = mean(lifeExpectancy))
print(LE_Mean_2019)
```

5.  (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see.\
    Hint: these data (data in long form) is not the simplest to compute growth. But you may want to check out the lag() function. And do not forget to group data by continent when using lag(), otherwise your results will be messed up! See [**https://faculty.washington.edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute**](https://faculty.washington.edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute).\

#### Asia has the largest growth in Life Expectancy while Europe has the lowest Life Expectancy in the year 2019.

```{r}

df %>% 
  filter(time %in% c("1960", "2019"), !is.na(lifeExpectancy), !is.na(region)) %>%
  group_by(region, time) %>% 
  summarize(avg_LE = mean(lifeExpectancy)) %>%
  mutate(last = lag(avg_LE), growth_le = avg_LE - last) %>%
  filter(!is.na(growth_le)) %>%
  arrange(desc(growth_le))


```

6.  (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both histograms on the same graph, see how well you can do it!

#### *This histogram shows that over time more countries are having more GPD per capita growth from the year 1960 to 2019. You can see this by the purple bar which represent the year 2016.*

```{r}
df %>%
  filter(time %in% c(1960, 2019), !is.na(GDP_PC)) %>%
  ggplot(aes(x = GDP_PC, fill = factor(time))) +
  geom_histogram(alpha = 0.9, position = "dodge", bins = 45) + 
  scale_fill_manual(values = c("orange", "purple"), labels = c("1960", "2019")) +
  labs(x = "GDP per capita", y = "Count", fill = "year") +
  theme_minimal()

```

7.  (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When counting from top.)\
    Hint: check out the function rank()!\
    Hint2: 17 for 1960.\

#### *For 1960 the rank of the United state is 17, while for 2019 the United state was 46. Meaning that from 1960 to 2019, the US dropped 29 ranks. This was probably due to better technology and medicine.*

```{r}

df %>% 
  filter(time == "1960", !is.na(lifeExpectancy), !is.na(name)) %>% 
  select(name, lifeExpectancy) %>%
  arrange(desc(lifeExpectancy)) %>%
  mutate(lifeExpectancy = rank(-lifeExpectancy)) %>% 
  filter(name == "United States of America") %>% 
  select(name, lifeExpectancy) %>% 
  print()

df %>% 
  filter(time == "2019", !is.na(lifeExpectancy), !is.na(name)) %>% 
  select(name, lifeExpectancy) %>%
  arrange(desc(lifeExpectancy)) %>%
  mutate(lifeExpectancy = rank(-lifeExpectancy)) %>% 
  filter(name == "United States of America") %>% 
  select(name, lifeExpectancy) %>% 
  print()

```

8.  (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019--what about the relative rank divided by the corresponding number of countries that have LE data in the corresponding year?\
    Hint: 0.0904 for 1960.\

#### *The LE for the United State has been dropping because over the course of the year, more countries has better technology and medicine to increase there LE. Which push the United down the rank.*

```{r}

df %>% 
  filter(!is.na(lifeExpectancy),time == "1960", !is.na(region)) %>%
    mutate(rank_num = rank(-lifeExpectancy), number_country = n(), relative_rank = rank_num/number_country) %>% 
    select(name, rank_num, number_country, relative_rank) %>% 
    filter(name == "United States of America") %>% 
    print() 

```

------------------------------------------------------------------------

Finally tell us how many hours did you spend on this PS.

It was roughly more than 20 hours + 3 hours from office hours.
